\chapter{Introduction}

%\begin{chapquote}{Author (\citeyear{})}
%``''
%\end{chapquote}

%\begin{displayquote}
%\textit{[T]he humid evening air is filled far and wide with the fragrance of the ripe ananas. The stalks of the pineapples, swelling with rich juice, rise between the lowly herbs of the meadow, and the golden fruit is seen shining at a distance from under its leafy crown of bluish-green.}

%\hspace*{\fill} \textbf{---}  Alexander von Humboldt
%\end{displayquote}



\section{Problem Statement}

Upon his return from his second voyage to America, Columbus presented King Ferdinand and Queen Isabella with gifts including gold nuggets, native artefacts, and various exotic birds, trees, animals, and plants, including a pineapple. Most pineapples rotted during the long voyage, but the king and queen tasted the one unspoiled and declared that they preferred it to all other fruits \citep{o2013pineapple}. Centuries later, pineapples have become a common import in temperate climates. 

In 2018, pineapple ranked 9th as the most-harvested fruit in the world, with a global production of $28 \times 10^6$ tonnes; the most harvested fruit in the world, banana, amounted to $114 \times 10^6$ tonnes \citep{fruit2020international}. The five leading producers of pineapple are Costa Rica, the Philippines, Brazil, Indonesia, and China. From these, Costa Rica is the largest exporter ($2.2 \times 10^6$ tonnes in 2021), and the second-largest producer, amounting to $2.9 \times 10^6$ tonnes in 2021 with an average increase of 6.7\% in the last 20 years \citep{FAOSTAT2022}. 

What once was one of the many riches sent from the colonies back to the mother countries is now a relevant income for exporters. The pineapple industry contributes approximately 1.7\% of Costa Rica's GDP, and it provides 32,000 direct jobs and 120,000 indirect jobs nationally \citep{chen2020production, employmentCR}. From the land surface used for permanent croplands in the country, 11\% is dedicated to pineapple, which tops the list next to coffee, palm oil, sugar cane, and banana \citep{cenagro2014}. 

In an attempt to grow the \textit{fruit of kings} in the Europe of the 17th century, hothouses were employed to maintain heat by ingenious means at enormous cost \citep{o2013pineapple}. Today, different technologies coupled with international trade provide easy and inexpensive access to the \textit{queens of fruits} and present high profitability to producing countries. Yet, regardless of the time and place, artificial methods of growing crops interfere with the natural ecosystem. Playing a relevant role in the agriculture and economy of Costa Rica, the pineapple sector has gained domestic and international attention due to its environmental impacts affecting the nation at large.

Grown as a monoculture, pineapple farming in Costa Rica presents the typical environmental problems of this type of cultivation practice, such as building up disease pressure, reducing particular nutrients in the soil, erosion due to conventional ploughing, and contamination of soil and water sources \citep{rodriguez2020agricultural, salaheen2019organic}. Moreover, the reduction of livestock productivity caused by the stable fly (\textit{Stomoxys calcitrans}) bred in the crop residues has become a major problem for pineapple producers and ranchers alike \citep{alpizar2016analisis, elbersen2019costa}. This cross-sectoral problem is difficult to tackle with a fruit that produces large amounts of crop residues. Farmers usually plant between 65,000 and 80,000 pineapples per hectare, and these can grow to weigh between 2.5 and 3.0 kg, with a height and width of between 1 and 2 metres \citep{asim2015review}. After the last harvest, these residues, also referred as stubble, are left in the field. 

%Pineapple stubble management is a challenging externality for a sector whose growth is associated with economic benefits on the one hand, and environmental concerns on the other \cite{poltronieri2016biotransformation}. 

Currently, to avoid the breeding of the stable fly in the stubble, farmers implement various practices, including drying with herbicides, burning, burying, and natural decomposition. These practices are both economically and environmentally costly. As \cite{hernandez2018impact} note, the costs of stubble management per hectare with these practices can range between US\$ 1,000 and US\$ 2,500 depending on what method is used. Moreover, these practices increase GHG emissions, hinder pineapple production productivity, and affect communities near the fields \citep{cesarino2020fabrication,netz2007climate}. Monitoring and controlling crop residue practices is not an easy task for an industry that has expanded in the country without regulation \citep{rodriguez2020agricultural}.

The pressing issue of stubble management and its consequences has incentivised the exploration of innovative solutions, mostly focused on the valorisation of pineapple leaves (PAL). these solutions not only help to reduce the financial costs and environmental damages of the current practices, but also contribute to the circular economy (CE) transition and to the bioeconomy. CE initiatives aim to move away from traditional ``make and dispose" models and towards systems that encourage material efficiency, and the objective of the bioeconomy is to use renewable biological resources sustainably to produce food, bioenergy and biobased goods. Thereby, these models help to create more circular material systems which reduce energy and emissions \citep{IPCC_2022_WGIII_SPM}. 

The valorisation of pineapple leaves is not new, their use for various purposes, including weaving, netting, and rope-making, has a long history among the natives of Amercia, preceding the arrival of Europeans \citep{collins1949history, o2013pineapple}. Today's linear production model, which leads to vast amounts of stubble produced in the pineapple industry, makes the implementation of valorisation solutions challenging. Different options to valorise the stubble have been tested in Costa Rica, e.g., producing fodder to feed livestock, or using the leaves of the plant to produce pineapple leaf fibre (PALF). Nevertheless, as of today, there is no large-scale stubble valorisation project in the country, and progress towards a systematic implementation seems stagnant. 

Despite the extensive tabletop research on technological solutions to produce biobased materials and bioenergy with PAL, socioeconomic studies on the implications of a Circular Bioeconomy (CB) transition are lacking. The adoption of innovative practices needed for this transition can be hindered by the existence of several bottlenecks, such as financial constraints, cultural and operational barriers, market demand uncertainty, and technological limitations. Potential solutions have been identified, but many questions related to their implementation are unanswered. Moreover, there seems to be no consent among stakeholders on the required steps for the valorisation process to take off. In this sense, an understanding of the actors, institutions, and the networks connecting them is needed. 

Finally, we must consider that many of the barriers preventing the transition to a CB are specific to Costa Rica, and solutions that have been implemented in other places may not apply to the case of Costa Rica. For example, the pineapple production clusters and the distinctive road network of the country requires finding tailored operational solutions. As of today, there is no vision of how a large-scale valorisation operation would take place in the industry. By providing realistic solutions that can be applied systematically, stakeholders can gain momentum to make progress towards a sustainable solution more easily. Ultimately, by understanding what is preventing the valorisation transition and by identifying actionable challenges, the pineapple industry in Costa Rica can advance in the agricultural CB revolution. 

\section{Objectives and research questions}
\label{researchQ}

Considering the knowledge gap identified within the problem statement, the present research considers the following aim and objectives. The aim is to help increase the sustainability of the pineapple industry by introducing circular bioeconomy principles. Two objectives have been defined for this aim: First, to explain why the valorisation of PAL has not taken off in the country and to explain the complexity of the system in which the valorisation unfolds. The second objective is to help to understand how a large-scale valorisation process could be carried out operationally. 

These objectives are defined given the current (early) stage at which the valorisation of PAL is in Costa Rica. Moreover, the absence of socioeconomic data and abundance of uncertainty led us to consider an exploratory, qualitative, case-based study for our analysis.

To attain the proposed objectives, several research questions and subquestions are proposed:

     
\begin{enumerate}

    \item A New Economy for An Old Problem
    \begin{itemize}
        \item What is the state-of-the-art technology for extracting PAL from the field, and what are its estimated costs?
        \item What are the valorisation options being developed in CR, and what is their development stage?
        \item What are the potential business models for the PAL?
        \item What is the demand for the potential PAL-based products?
        \item What local regulations and characteristics should be considered when implementing valorisation options in CR?
    \end{itemize}
    
    \item Harvesting The Fruits Of Uncertainty
    \begin{itemize}
        \item What are the cultural, financial, market-related, operational, and technological barriers preventing the valorisation of pineapple stubble in Costa Rica?
        \item What is needed to overcome these barriers? Whose action is required?
        \item What are the benefits and challenges of valorising the stubble?
    \end{itemize}
    
    \item The Pineapple Leaves Route
    \begin{itemize}
         \item What are suitable locations for PAL processing plants?
        \item What is the optimal spatial distribution of PAL processing plants? 
        \item Should the processing of PAL be centralised or decentralised? 
     \end{itemize}
    
\end{enumerate}

\section{Overview of the structure of the thesis}

The valorisation of PAL and its development in Costa Rica are explored from different angles. Thus, this paper has been divided into three chapters, following the same order as the research questions delineated abovde.

In \cref{neweconomy}, we provide a review of the pineapple stublle and its valorisation within the Costa Rican context. We first give a brief explanation of the current management of pineapple stubble in the field. We then discuss the extraction of PAL from the field and its singularities. Then, the potential valorisation options and demand for PAL are explained. Finally, a brief discussion of the local context and regulations applicable to the valorisation of PAL is provided.

\cref{chapter_3lab} explains the study of the barriers preventing the valorisation of PAL in Costa Rica. We first provide appropriate theory related to circular (bio)economy and circular-orientes innovation. We continue by explaining the methodology employed for the qualitative analysis, Fuzzy Cognitive Mapping. Then, results of the study are shown and discussed. Finally, conclusions and recommendations are provided. 

The use of location analysis can be useful to unravel the operational challenges of valorising PAL. This analysis is explained in \cref{FLPchapt4}. We first introduce the Facility Location Problem. Then, we provide an explanation of the methodology used for the case study. Results are depicted, followed by a discussion. Conclusion and recommendations are provided. 

The scope and aim of this study have evolved throughout its development. Also, the author's perception and understanding of the issue at hand have undergone a transformation since the beginning of the research. Thus, in \cref{concludeGen}, general conclusions of the study are given, and a reflection from the author regarding the topic of study and the employed methods finalises the paper. 

