\begin{footnotesize} 
\begin{Parallel}[c]{0.48\textwidth}{0.48\textwidth}

\bi{\centering\textit{A Colón}}{\centering\textit{To Columbus}}

\vspace{5pt}

\bi{¡Desgraciado Almirante! Tu pobre América, \\
tu india virgen y hermosa de sangre cálida, \\
la perla de tus sueños, es una histérica \\
de convulsivos nervios y frente pálida. }{Unfortunate admiral! Your poor America, \\
your beautiful, hot-blooded, virgin Indian love, \\
the pearl of your dreams, is now hysterical, \\
her nerves convulsing and her forehead pale.}

\vspace{5pt}

\bi{Un desastroso espirítu posee tu tierra: \\
donde la tribu unida blandió sus mazas, \\
hoy se enciende entre hermanos perpetua guerra, \\
se hieren y destrozan las mismas razas.}{
A most disastrous spirit rules your land: \\
where once the tribesmen raised their clubs together, \\
now there is endless warfare between brothers, \\
the self-same races wound and destroy each other.}

\vspace{5pt}

\bi{Al ídolo de piedra reemplaza ahora \\
el ídolo de carne que se entroniza, \\
y cada día alumbra la blanca aurora \\
en los campos fraternos sangre y ceniza.}{The stone idol is gone, and in its place \\
a living idol sits upon a throne, \\
while every day the pallid dawn reveals \\
the blood and ashes in the fields of neighbours.}

\vspace{5pt}

\bi{Desdeñando a los reyes nos dimos leyes \\
al son de los cañones y los clarines, \\
y hoy al favor siniestro de negros reyes \\
fraternizan los Judas con los Caínes.}{
Disdaining kings, we give ourselves our laws \\
to the sound of cannons and of bugle-calls, \\
and now, on the sinister behalf of black kings, \\
each Judas is a friend of every Cain.}

\vspace{5pt}

\bi{Bebiendo la esparcida savia francesa \\
con nuestra boca indígena semiespañola, \\
día a día cantamos la Marsellesa \\
para acabar danzando la Carmañola.}{
We love to drink the festive wines of France; \\
day after day we sing the Marseillaise \\
in our indigenous, semi-Spanish voices, \\
but end by roaring out the Carmañola.}

\vspace{5pt}

\bi{Las ambiciones pérfidas no tienen diques, \\
soñadas libertades yacen deshechas. \\
¡Eso no hicieron nunca nuestros caciques, \\
a quienes las montañas daban las flechas!}{
The treacheries of ambition never cease, \\
the dream of freedom lies in broken bits. \\
This crime was never committed by our chiefs,\\
by those to whom the mountains gave their arrows.}

\vspace{5pt}

\bi{Ellos eran soberbios, leales y francos, \\
ceñidas las cabezas de raras plumas; \\
¡ojalá hubieran sido los hombres blancos \\
como los Atahualpas y Moctezumas!}{
They were majestic, loyal, and great-hearted; \\
their heads were decorated with rare feathers. \\
Oh if the white men who came had only been \\
like the Atahualpas and the Moctezumas! }

\vspace{5pt}

\bi{Cuando en vientres de América cayó semilla \\
de la raza de hierro que fue de España, \\
mezcló su fuerza heroica la gran Castilla \\
con la fuerza del indio de la montaña.}{When once the seed of the iron race from Spain \\
was planted in the womb of the Americas, \\
the heroic strength of great Castile was mixed \\
with the strength of our own Indians of the mountains.}

\vspace{5pt}

\bi{¡Pluguiera a Dios las aguas antes intactas \\
no reflejaran nunca las blancas velas; \\
ni vieran las estrellas estupefactas \\
arribar a la orilla tus carabelas!}{Would to God that these waters, once untouched, \\
had never mirrored the white of Spanish sails, \\
and that the astonished stars had never seen \\
those caravels arriving at our shores!}

\vspace{5pt}

\bi{Libre como las águilas, vieran los montes \\
pasar los aborígenes por los boscajes, \\
persiguiendo los pumas y los bisontes \\
con el dardo certero de sus carcajes.}{The mountains saw how the natives, who were free \\
as eagles, came and went in the wild forest, \\
hunting the deer, the puma, and the bison \\
with the sure arrows they carried in their quivers.}

\vspace{5pt}

\bi{Que más valiera el jefe rudo y bizarro \\
que el soldado que en fango sus glorias finca, \\
que ha hecho gemir al zipa bajo su carro \\
o temblar las heladas momias del Inca. }{A chief, though rough and bizarre, is worth far more \\
than a soldier who roots his glory in the mud, \\
who has caused the brave to groan beneath his car \\
or the frozen mummies of Incan lords to tremble.}

\vspace{5pt}

\bi{La cruz que nos llevaste padece mengua; \\
y tras encanalladas revoluciones, \\
la canalla escritora mancha la lengua \\
que escribieron Cervantes y Calderones. }{The cross you brought to us is now decayed,\\
and after the revolution of the rabble,\\
the rabble writing today defiles the language\\
written by great Cervantes and Calderon.}

\vspace{5pt}

\bi{Cristo va por las calles flaco y enclenque, \\
Barrabás tiene esclavos y charreteras, \\
y en las tierras de Chibcha, Cuzco y Palenque \\
han visto engalonadas a las panteras. }{A gaunt and feeble Christ walks through the streets,\\
Barrabas can boast of slaves and epaulets,\\
and the lands of Chibcha, Cuzco, and Palenque\\
have seen wild beasts acclaimed and decorated.}

\vspace{5pt}

\bi{Duelos, espantos, guerras, fiebre constante \\
en nuestra senda ha puesto la suerte triste: \\
¡Cristóforo Colombo, pobre Almirante, \\
ruega a Dios por el mundo que descubriste!}{Evil mischance has placed afflictions, horrors,\\
wars, and unending fevers in our way:\\
Oh Christopher Columbus, unfortunate admiral,\\
pray to God for the world that you discovered!
}


\renewcommand{\ParallelAtEnd}{\hfill\textit{Ruben Darío (1867--1916)}
}

\end{Parallel}

\end{footnotesize}

